---
title: "00_HelloR"
output: html_document
---

```{r setup, include=FALSE}

```

# Basics of R

In this activity we will cover what R is, why is it so popular, and what it can do for us.

We will learn about the R Studio User Interface. 

Creating a project.

Running R commands through interactive shell.

Saving R commands in Scripts.

R as a programming language.

# R as a Programming Language

## R as an Overengineered Calculator

```{r}
2+2
2*3
2^5
```

## Variables

### Scalar

```{r}
A <- 2
A
B <- 3
B
A * B
```
### Array

```{r}
C <- 2:11
C
D <- c(2, 4, 6)
D
A*D
```

Indexing

```{r}
D[2]
C[2:4]
C[D]
```

### Dataframe

```{r}
data(mtcars)
mtcars
```

Indexing

```{r}
mtcars[1,]
mtcars[,3:5]
mtcars[,"disp"]
mtcars$disp

mtcars[1:5, c("mpg","disp")]
```

## Functions

Existing functions

```{r}
help()
```

Some functions are argumentative.

```{r}
help(min)
min(mtcars$disp)
mean(mtcars$disp)
max(mtcars$disp)
```

Some functions are more argumentative.

```{r}
lm(mpg ~ disp + am, mtcars)
```

You can also write your own functions, but 99% of the time its better you use existing functions.

```{r}
Hello <- function(x){
  print(paste("Hello",x))
}
Hello("Mesterskaberne")
```