---
title: "Working Environment"
output: html_document
---

```{r setup, include=FALSE}

```

# Packages

So if its better to use existing functions 99% of the time. Where does one get them?

Answer is packages. R is an open source project and people develop packages for it all the time.

As of February 2022, there are over 18000 packages in CRAN.

Easier way to handle packages is to use the GUI. Before a package can be used:

+ Its files need to be installed
+ It should be loaded to the memory

Below is how to do it from the command line.

```{r}
# Install files needed
install.packages("ggplot2")
# Load it to memory
library(ggplot2)
```

# Working Directories

When you launch R, the folder you launched R from becomes the working directory.

You can refer to the files in the wd with relative paths, the files you save will be saved here.

On GUI, you can do it in the files tab.

Here is how to do it from command line.

```{r}
# Find out what our wd is
getwd()
# Save wd in a variable
myWD <- getwd()
# Here is how you change wd. In this case it won't be changed.
setwd(myWD)
```
# Projects

R studio is a really useful IDE. The work you do in R Studio will be saved as a project.

This allows you to pick up quickly where you left off. 