---
title: "Descriptives"
output: html_document
---

```{r setup, include=FALSE}
# Import Data
library(ggplot2)
data(mpg)
# To get away from tidyverse
mpg <- data.frame(mpg)
```

In this activity we will learn about taking a quick look at the data once it is 
imported. We will use the mpg dataset included in ggplot2 package.

## Structure

Once the data is in, first thing you will want to know is how many variables, 
and how many observations are in it. Followed by what type of variables are 
included.

Use str function to display the structure of the dataframe.

```{r}
str(mpg)
```
First line tells you there are 32 observations (rows) and 11 variables (columns).

It is followed by a list of columns. The chr, num, int tell you the type of data.

Then you see the first few values of the variable.

The type of data is very important as this tells you what to expect and can
inform your decisions to visualize and model this data.

## Descriptives

Next you will want to know is the descriptive statistics on the data. 

What is the average, maximum, minimum and so on. This information can give you an
inkling of what the distribution looks like.

```{r}
summary(mpg)
```

You can see the information we mentioned in the output. There are no missing values
in this dataset. If there were, we would also find out about them here.

## Individual Functions

Sometimes you may want to refer to descriptives on their own, or use them in your
calculations. There are individual functions for this type of information.

Here are some examples.

```{r}
# Observations
nrow(mpg)
# Variables
ncol(mpg)
# Average city mpg
mean(mpg$cty)
# Max highway mpg
max(mpg$hwy)
# min number of cylinders
min(mpg$cyl)
# standard deviation of hwy
sd(mpg$hwy)
```