---
title: "02_Visualizations"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
data(mpg)
load("Profits.RData")
```

# Visualizations 

Visualizations serve two purposes.

+ Explore Data
+ Communicate information

In R you can either use the base R functionality or the newer ggplot2 package.

ggplot2 is very flexible and powerful. You can visualize anything with it.

The package is based on grammar of visualization. Each visualization is a combination of:

+ Data
+ Aesthetics
+ Geometric Objects

## GGPLOT - Syntax

Data is mpg

Aesthetics is we assign displacement to x axis and cty to y axis.

```{r}
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point()
```

qplot is a wrapper around ggplot that eases its use. It uses sane defaults.

```{r fig.width=21,fig.height=9}
qplot(x = displ, y = cty, data = mpg)
```

So if what you want is run of the mill visualizations qplot can be a good solution.

## Basic Visualizations

Here are some basic visualizations:

+ Histograms
+ Barcharts
+ Scatter Plots
+ Line Plots 
+ More...

We will see when to use which of these.

### Histogram

Single continuous variable.

```{r}
ggplot(mpg, aes(x = cty)) +
  geom_bar()
```

### Barchart

Single categorical variable.

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
```

### Scatter Plot

Relation between two variables.

```{r fig.width = 21, fig.height = 9}
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point()
```

### Line Plot

Variables that change over time.

```{r fig.width = 21, fig.height = 6}
ggplot(profits, aes(x = year, y = profit, linetype = company, color = company)) +
  geom_line()
```

### Quick and Dirty Exploration

```{r}
plot(mpg)
```

## Making Things Clearer

Title's help

```{r fig.width = 21, fig.height = 6}
ggplot(profits, aes(x = year, y = profit, linetype = company, color = company)) +
  geom_line() +
  labs(Title = "Company A is more profitable")
```

Axis labels help

```{r fig.width = 21, fig.height = 6}
ggplot(profits, aes(x = year, y = profit, linetype = company, color = company)) +
  geom_line() +
  labs(Title = "Company A is more profitable",
       x = "Years since 0 AD",
       y = "Yearly profits reported in balance sheets")
```

Axis labels help

```{r fig.width = 21, fig.height = 6}
ggplot(profits, aes(x = year, y = profit, linetype = company, color = company)) +
  geom_line() +
  labs(Title = "Company A is more profitable",
       x = "Years since 0 AD",
       y = "Yearly profits reported in balance sheets") +
  ylim(0,120) +
  theme_bw()
```

To compare based on categories.

```{r fig.width = 21, fig.height = 6}
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point() +
  facet_wrap(~class) + 
  theme_bw()
```

## Saving Plots

You can use ggsave after you create a plot to save it.

```{r}
ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point() +
  facet_wrap(~class) + 
  theme_bw()
ggsave("FacetedPlot.png")
```