---
title: "Subsets"
output: html_document
---

```{r setup, include=FALSE}
zillow <- read.csv("zillow.csv")
```


## Introducing Data

Zillow is the boligsiden of USA. This dataset shows zillow rent index at zip code
level.

```{r}
zillow
```

## Subsetting with Indexes

Subsetting is selecting a smaller sample from available data, be it observations
(rows) or variables (columns). You may for example only want observations from
accounting department, or only people over a certain age... This is when 
subsetting becomes useful.

### Variables

You can select a column by its column number. I will use head function to limit 
the output.

```{r}
head(zillow[, 5])
```
You can use the name of the column.

```{r}
head(zillow[, "Metro"])
```
You can select multiple columns by column number.

```{r}
c(1:4, 8)
head(zillow[,c(1:4, 8)])
```

**Dropping a Column**

Let us say I want to drop a single column (MoM).

```{r}
colnames(zillow) # names of columns
# MoM is the 9th column
zillowSS0 <- zillow[, -9] # A negative index will exclude that column
colnames(zillowSS0)
```

You can also drop multiple columns by combining negative indices.

```{r}
zillowSS1 <- zillow[, c(-9, -10)]
colnames(zillowSS1)
```

I will now remove these subset datasets as they are cluttering the working 
environment (and to demonstrate rm() function).

```{r}
rm(zillowSS0, zillowSS1)
```

### Observations

You may be interested in only certain rows in a dataset. 

Let us say we want to focus on rent in Ohio counties only. 
Remember the logic operators from earlier learning activities. 

```{r}
zillowOH <- zillow[zillow$State == "OH", ]
```

zillow$State == "OH" will return an array of True/False results. 
It will be true where State is Ohio, and false otherwise. When we feed this 
logical operation into the row index, and R will only return observations where 
evaluation resulted in True.

We can also combine multiple criteria in our filtering. Let us say we are 
interested in rent in Ohio counties but only for areas where Zri is lower than $1000.

```{r}
zillowOHCheap <- zillow[zillow$State == "OH" & zillow$Zri < 1000, ]
```

We used an & (and operator) to combine two logical operations, R only returns 
results where both conditions evaluate True. 

Let us say you have some flexibility, you are looking for a rental in Ohio 
***or*** West Virginia.

```{r}
zillowOHWV <- zillow[zillow$State == "OH" | zillow$State == "WV", ]
```

Here we used a | (or operator) to combine two logical operations, R only returns 
results where at least one condition is True.

Let us clean the workspace by removing some datasets.

```{r}
# Yeah, this may be a bit obscure at the moment. Try to figure it out, its good exercise.
rm(list = ls(pattern = "zillowOH*"))
```

Here is a little exercise for you. Considering there are `r nrow(zillow)` 
observations in zillow dataset, can you make R randomly select 50 numbers 
between 1 and `r nrow(zillow)` and create a random subset? 

Hint: read about sample function.

## Subset Function

Subset function can do both variable and column subsetting at the same time and it has a more intuitive interface than indexing.

Check the manual pages. 

```{r}
?subset
```

The subset syntax is as follows:

    subset(x, subset, select, drop)
    
  + x is the data to be subsetted.
  + subset is the criteria for rows.
  + select is the list of columns to keep
  
only x is mandatory.

### Selecting Variables

Let us just select State, County, and Zri from zillow dataset.

```{r}
# Note the lack of quotation marks
zillowSS0 <- subset(zillow, select = c(State, County, Zri))
colnames(zillowSS0) # names of columns
head(zillowSS0) # A sampling of data
```

Here I called subset with two parameters.
1. zillow: I specified which data is to be subsetted.
2. select=c(State, County, Zri): I specified which columns I wanted to keep.

One nice thing about subset is that you can use ranges of columns with names 
just like with ranges of numbers. Let us say we want all variables between date and zri.

```{r}
# Note the lack of quotation marks
zillowSS1 <- subset(zillow, select = Date:Zri)
colnames(zillowSS1) # names of columns
head(zillowSS1) # A sampling of data
```

Here I called subset with two parameters again.
1. zillow: the dataset
2. select=Date:Zri: I specified a range of columns from Date to Zri.

### Selecting Observations

Just as in the indexing, we can use logical operators to extract specific rows.

Let us carry out that example of Ohio or West Virginia.

```{r}
zillowOHWV <- subset(zillow, State == "OH" | State == "WV")
head(zillowOHWV)
```

Here I called subset with two parameters
1. zillow: the dataset
2. State == "OH" | State == "WV": State is OH or WV 

Plain language translation of this command is: "get me the rows from zillow 
dataset where the state is either Ohio or West Virginia."

```{r}
rm(zillowOHWV)
```

### Selecting Both

We can subset both variables and observations at the same time.

```{r}
zillowOW <- subset(zillow, 
                   (State == "OH" | State == "WV") & Zri < 1000,
                   select = Date:Zri)
```

Same idea, only made more complicated by including both row and column subsetting. 
Parameters:

1. zillow: dataset
2. (State=="OH" | State=="WV") & Zri<1000: State is Ohio or West Virginia, and Zri is less than 1000
3. select=Date:Zri: columns between Date and Zri

Plain language translation of this command is: "Get me columns from Date to Zri 
of the zillow dataset, where the state is either Ohio or West Virginia, and the Zri is less than 1000."

```{r}
rm(zillowOW)
```
