---
title: "Assignment 2"
author: "Irfan Kanat"
date: "March 8, 2018"
output:
  html_document: default
  pdf_document: default
---
This is an assignment to evaluate what we have learned in module two. For each learning activity there is a section in this document. Answer each question below the question and create an R Markdown document for submission.

*For the rest of the exercises we will use games.csv in the data folder. Import the file into a data.frame called games to begin the exercise.*

```{r setup}
# Import necessary packages

# Import games.csv from data folder

```

## 1 - Describe

### Visual Inspection

Show the top 6 rows of *games*.

```{r}

```

### Descriptive Statistics

Use the *summary* function to provide basic descriptive statistics for *games*.

```{r}

```

Use the *str* function to provide an overview of data structure for *games*.

```{r}

```

## 2 - Visualization

### Histogram

Plot out the distribution of price variable in a histogram.

```{r}

```

### Scatter Plot

Plot the relation between avgMin (on average how much time players spent in game) and reviews.

```{r}

```


### Bar Chart

Plot the frequency of status types in *games*.

```{r}

```

## 3 - Subsetting

### Selecting Rows and Columns

Show only the games that have SuperStar status.

```{r}

```

Show only the first three rows.

```{r}

```

### Aggregating

Calculate the average price for each status category, save the results into a data.frame called *gameAvgPrice*.

```{r}
# Using formula interface for aggregate

```

Change the name of the column listPrice in *gameAvgPrice* to avgCategoryPrice.

```{r}

```

### Combine Datasets

Combine *games* and *gameAvgPrice* so that the average price for category is displayed next to each game. Save the results into *games* data.frame.

I want column avgCategoryPrice to be added to games, in such a way that it shows the average price for the correct status. A game that is a niche player will show 10.87 in avgCategoryPrice for example.

```{r}

```

## Transformations

Add a new column to *games* named logListPrice, and store log transformed list price in this column. Don't forget log of 0 is meaningless, take the necessary precautions to ensure this does not lead to loss of data.

```{r}

```

## Missing Data

Show how many rows there are in games.

```{r}

```

Exclude any observations that are missing data on reviews column. Save the resulting data.frame in *games*.

```{r}

```

Show how many rows there are after removal of observations.

```{r}

```

## Dimension Reduction

WIP
